Fourier-domain modulations and delays of gravitational-wave signals

### 2018-06-18 docs/paper_method.tex
Edits to the paper.

We present a Fourier-domain based approach modulations and delays of gravitational wave signals, a problem which arises in two different contexts. For space-based detectors like LISA, the orbital motion of the detector introduces a time-dependency in the response of the detector, consisting of both a modulation and a varying delay. In the context of signals from precessing spinning binary systems, a useful tool for building models of the waveform consists in representing the signal as a time-dependent rotation of a quasi-non-precessing waveform. In both cases, being able to compute transfer functions for these effects directly in the Fourier domain may enable performance improvements for data analysis applications by using fast frequency-domain waveforms. Our results generalize previous approaches based on the stationary phase approximation, extending them by including delays and higher-order corrections while being applicable to a broader class of signals including full inspiral-merger-ringdown waveforms. In the LISA case, we find that higher-order corrections remain small except at the two extremes of the detector's frequency band, where we show that including them reduces the errors. In the case of precessing binaries, we perform a limited exploration of the merger-ringdown range based on a toy model for the precession. We find that, depending on the post-merger frame velocity, higher-order corrections can become quantitatively large to the point of challenging our perturbative formalism, while affecting mainly the amplitude and giving rise to limited unfaithfulness. We further present an alternative direct convolution approach to accurately represent these post-merger features. \SM{[update to include SOBHs]}

@@

\jgb{The GW community is making progress assembling higher-fidelity tools for these coming challenges.
State-of-the-art IMR templates combine information from the perturbative results of post-Newtonian theory covering the inspiral (see e.g.~\cite{BlanchetLiving}) and from numerical relativity simulations covering the end of the inspiral and the merger-ringdown phase (see e.g.~\cite{Pfeiffer12}).  Approaches to template construction include phenomenological templates postulating an analytic ansatz for the Fourier-domain amplitude and phase~\cite{Husa+15, Khan+15,Hannam+13}, and the more physically motived Effective-One-Body (EOB) approach~\cite{BD99,Taracchini+13, Pan+13, Bohe+16}. Where needed, Reduced Order Models (ROM), also called surrogate models, have been developed to considerably speed up waveform generation, without losing accuracy~\cite{Field+13, Puerrer14, Bohe+16}. These tools provide efficient non-precessing IMR Fourier-domain waveforms. Importantly for this work, the resulting waveforms can be represented by an amplitude and phase for each mode, with only a few hundred samples~\cite{Puerrer14}.[JB:I cut this down significantly.]} Our challenge will be to accurately complete the representation of instrumental signals including precession effects and the LISA instrumental response and precession effects without sacrificing the computational efficiency enabled by these fast waveform templates.

A complete representation of spin effects in fast IMR templates still remains a frontier of gravitational wave signal modelling. In presence of misaligned spins, the system will endure precession of the orbital plane as it evolves, leading to modulations of the signal as seen by the observer~\cite{ACST94}.
%Precessing EOB templates incorporate all six parameters of the two spins~\cite{Pan+13, BTB16} but still require a costly time-domain integration; while precessing phenomenological templates~\cite{Hannam+13} decrease the dimensionnality by using effective spin parameters representing only the dominant effects of the spins.
\jgb{[JB:I cut this down too.]}

A promising approach to modeling the effect of precession on the emitted waveform, it has been proposed~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11} to decompose precessing waveforms by performing a time-dependent rotation, following the precession of the orbital plane. The resulting waveform in the rotated frame can then be modelled by a non-precessing waveform, an approximation which is used both in the construction of the inspiral part of precessing EOB waveforms~\cite{Pan+13} and in the construction of precessing phenomenological waveforms~\cite{Hannam+13}. To follow this modelling approach and efficiently create Fourier-domain waveforms, one needs to understand how to translate the time-domain modulations created by the frame rotation into a Fourier-domain transfer function, which is our first motivation.

Beyond a fast representation of the incident waveform, observational analyses also require transforming the signals through some instrumental response. For short duration mergers such as LIGO  has detected, this can be treated by a simple multiplier.  For future instruments though, the instrumental response will be more complicated.

@@

\jgb{Whereas LIGO and Virgo are typically sensitive to chirping binaries for a minute or less, LISA signals may accumulating over months or years.
The response of a LISA-type instrument is thus time-dependent~\cite{Cutler97}.
The motion and change of orientation of the detector constellation along its orbit lead to significant time variability in the form of a modulation and a varying delay. These effects then convey information about the localization of the gravitational-wave source in the sky. Direct implementation of the detector response is straightforward\cite{SynthLISA,LISACode,LISASim} but at prohibitive computational cost for parameter-estimation analyses. To leverage the performance of state-of-the-art Fourier-domain IMR templates~\cite{BTB16,Khan+15}, we must efficiently process the signals through the time-dependent response of the dectector while staying in the Fourier domain.}

@@

\jgb{In the case of precessing binaries, applying the SPA directly to the modulated signal is also prone to pathologies. In Ref.~\cite{KCY13,KCY14}, a formalism (called shifted uniform asymptotics or SUA) was introduced to go beyond the SPA and compute more accurately the modulation in the Fourier domain; however, this formalism still relies on the SPA for the underlying precessing-frame signal, and is as such limited to inspiraling signals. \textbf{[JB: Did I capture your thoughts correctly in the following.]}
The simplified treatment of the precession response in the phenomenological waveforms in Ref.~\cite{Hannam+13} takes another approach, treating the precession modulation in the frequency domain by directly associating the Fourier frequency with (a multiple of) the post-Newtonain orbital frequency. As we will explain, this corresponds to the zeroth-order approximation of the SUA.}

In the case of the LISA response, the SPA provides a natural map from time-domain (for the orbit) to Fourier-domain (for the signal)~\cite{Cutler97}. Ref.~\cite{Klein+15} included the orbital motion of the detector in the SUA treatment. Consistently extending these previous approaches to the merger-ringdown part of the signals, including the delays in the LISA case, is part of the objectives of this paper.

We seek to overcome two limitations in such previous approaches.  The first is that  SPA-based methods are not applicable to IMR waveforms. Second, there is often no clear way to improve the accuracy of these methods beyond the intuitive leading order treatment in order to meet the high-accuracy needs of future detectors. Our approach exploits separation of time-scales approximations, based on a general treatment directly in the Fourier domain of slowly varying delays and amplitude modulations for chirping waveforms with slowly varying amplitude and phase.

The plan of the paper is as follows. In Sec.~\ref{sec:motivation}, we provide a general presentation of problem of Fourier-domain modulation and introduce the relevant timescales for both the response of LISA-type detectors and the modulation of precessing signals. In Sec.~\ref{sec:formalism}, we present our general formalism, give its leading order approximation as well as higher-order corrections, introduce new timescales based on the Fourier-domain signal, refine the previous results for both the quadratic-in-phase corrections and the treatment of the delays, and make the connection to previous approaches of the problem. We then apply our formalism to the response of the LISA detector in Sec.~\ref{sec:LISA}, and to the case of signals from precessing binaries in Sec~\ref{sec:precession}. We discuss and summarize our results in Sec.~\ref{sec:discussion}.

\SM{[say that we reverse point of view: waveform is given FD, modulation TD, instead of waveform+modulation TD]}

\SM{[say a word of how to gauge relevance of a given error: 1/SNR in direct error ? We won't give anything but waveform reconstruction errors...]}

\SM{[check citation of Cornish-Rubbo-Poujade early work]}

@@

To understand the separation of timescales in the problem, it will be useful to have at hand the leading-order scaling laws for an inspiral (labeled the Newtonian order in the PN language). Although inaccurate for the purpose of waveform modelling, these leading-order estimates will give useful orders of magnitude of the relevant timescales in the inspiral. For a binary with masses $m_{1}, m_{2}$, we define the total mass $M=m_{1}+m_{2}$ and the symmetric mass ratio $\nu = m_{1}m_{2}/M^{2}$. Introducing a time of coalescence $t_{c}$, the relations between the orbital frequency and phase and the time to coalescence $t_{c} - t$ are then given by
\begin{subequations}\label{eq:omegaphiN}
\begin{align}
	\omega(t) &= \left[ \frac{256\nu}{5c^{5}} (GM)^{5/3} (t_{c}-t) \right]^{-3/8} \,, \\
	\varphi(t) &= -\left[ \frac{c^{3}}{5 G M \nu^{3/5}} (t_{c}-t) \right]^{5/8} \,.
\end{align}
\end{subequations}
As for the leading-order time-domain amplitude of the $22$ mode, we have~\cite{BlanchetLiving}
\be\label{eq:a22N}
	a_{22}^{\rm N} (t) = \frac{2 G M \nu v^{2}}{D c^{2}} \sqrt{\frac{16 \pi}{5}} \,,
\ee
where we set $v = (G M\omega/c^{3})^{1/3}$ and where $D$ is the luminosity distance to the observer. For this Newtonian inspiral, applying the SPA gives for the $22$ mode $\tilde{h}(f) = A_{\rm N}(f)e^{-i\Psi_{\rm N}(f)}$ with \SM{[check that total mass convention is $M$ everywhere]}
\begin{subequations}\label{eq:SPAN}
\begin{align}
	A_{N}(f) &= \frac{G^{2}M^{2} \pi}{Dc^{5}} \sqrt{\frac{2\nu}{3}} v^{-7/2}\,, \label{eq:ASPAN}\\
	\Psi_{\rm SPA}^{\rm N}(f) &= \phi_{0} - 2\pi f t_{0} - \frac{3}{128\nu v^{5}} \,, \label{eq:PsiSPAN} 
\end{align}
\end{subequations}
where $v=(G M \pi f/c^{3})^{1/3}$ according to the SPA correspondence~\eqref{eq:deftfSPA}, and where $t_{0}, \phi_{0}$ are constants. It is also customary to rewrite the above relations in terms of the chirp mass $\Mchirp \equiv M\nu^{3/5}$, which is the only mass combination characterizing the signal at the leading PN order.

@@

\SM{[explain somewhere the fact that the modulation has support for negative f, convolution has support to larger f - pretty important for direct convolution]}

\SM{[cite GWFrames]}

\SM{[replace $pp$ everywhere with $\perp\perp$]}

\SM{[check error estimates with single-spin configurations $+$, $\perp$, $-$]}

@@

An extension is required in specifying a model for the precession post-merger. After the merger occurs, the intuitive interpretation of the precessing frame as roughly following the orbital plane of the binary is lost. However, the prescription of~\cite{OShaughnessy+11} to extract the radiation axis is based entirely on the waveform itself, and can be applied to the ringdown as well. For $\ell=2$, analyzing numerical relativity waveforms, Ref.~\cite{OShaughnessy+12} described a qualitative model where the radiation axis essentially keeps precessing around the final angular momentum $\bm{J}$, but transitions to a faster precession rate, with an angular velocity determined by the quasinormal mode (QNM) frequencies of the remnant black hole as
\be\label{eq:OmegaframeQNM}
	\Omega_{\rm frame} = \omega_{220}^{\rm QNM} - \omega_{210}^{\rm QNM} \,.
\ee
This implies a significant acceleration of the precession post-merger with respect to the inspiral (see Fig.~\ref{fig:precmodel}). For the opening angle of the precessing cone $\beta$, Ref.~\cite{OShaughnessy+12} proposes an exponential decay driven similarly by the difference between damping frequencies. We found by inspection of various precessing waveforms currently available in the SXS catalog (see e.g.~\cite{SXScatalog, Mroue+12, Mroue+13}) that this picture is at least qualitatively correct for $\ell=2$, with $\beta$ appearing to transition to a lower value rather than decaying all the way to zero. By contrast, we found that the current ringdown prescription in the SEOBNRv3 model can cause a qualitatively different behaviour, with oscillations in $\beta$. \SM{[motivate this with final spin estimate as well, if we update this ?]}
%and its estimate for the spin of the final remnant~\cite{} would require to be updated (see e.g.~\cite{}) as shown in Fig.~\ref{fig:precmodel}.

@@

Here, we will use directly the wave polarizations $h_{+}$, $h_{\times}$, optimizing over time, phase, and polarization angle. For real or complex signals $a$, $b$, one introduces the usual Hermitian noise-weighted scalar product~\cite{CF94}
\be\label{eq:defoverlap}
	\left( a | b \right) = 2 \int \ud f \frac{\tilde{a}_{1}(f) \tilde{b}_{2}(f)^{*}}{S_{n}(f)} \,,
\ee
where the integral extends over all frequencies (in practice, two intervals $[-f_{\rm max}, -f_{\rm min}]$ and $[f_{\rm min}, f_{\rm max}]$). In the above, $S_{n}(f)$ the noise power spectral density, for which we use the aLIGO ZDHP noise curve~\cite{LIGOProspects13}. We use the notation $h \equiv h_{+} - i h_{\times}$ for the complex strain, and write $h[\delta t, \delta \Phi, \delta \psi]$ to represent a signal where we shifted the time of coalescence, the phase of the line of sight to the observer and the polarization angle. The dependencies in time and polarization are given by~\eqref{eq:shifttime} and $h[\delta \psi] = e^{2 i \delta \psi} h$, while the phase $\delta \Phi$ affects differently each waveform mode in~\eqref{eq:defmodes}. In the norm $(h|h)$, only the dependence on $\delta\Phi$ remains. For two signals $h_{1}$ and $h_{2}$, the mismatch between them is then defined as \SM{[check]}
\be\label{eq:defMM}
	\mathrm{MM} \equiv \mathrm{Min}_{\delta t, \delta \varphi, \delta \psi} \left( 1 - \frac{ \mathrm{Re} \left[ (h_{1}[\delta t, \delta \Phi, \delta \psi] | h_{2}) \right]}{\sqrt{(h_{1}|h_{1})[\delta\Phi]}\sqrt{(h_{2}|h_{2})}} \right)
\ee
where an optimization over a shift in time, phase and polarization is performed. This definition is the same as in~\cite{Blackman+17a} (see App.~D there), except that we added the optimization over the phase. The target unfaithfulness for waveform models depends on the application. For ground-based detectors like advanced LIGO and Virgo, it is often set to 1\% or 3\%, while detectors of the next generations, including LISA, will require a lower unfaithfulness for their higher SNR \SM{[acronym]} signals.

>>>>

We present a Fourier-domain based approach modulations and delays of gravitational wave signals, a problem which arises in two different contexts. For space-based detectors like LISA, the orbital motion of the detector introduces a time-dependency in the response of the detector, consisting of both a modulation and a varying delay. In the context of signals from precessing spinning binary systems, a useful tool for building models of the waveform consists in representing the signal as a time-dependent rotation of a quasi-non-precessing waveform. In both cases, being able to compute transfer functions for these effects directly in the Fourier domain may enable performance improvements for data analysis applications by using fast frequency-domain waveforms. Our results generalize previous approaches based on the stationary phase approximation for inspiral signals, extending them by including delays and computing corrections beyond the leading order, while being applicable to the broader class of signals of inspiral-merger-ringdown waveforms. In the LISA case, we find that a leading-order treatment is accurate for high-mass and low-mass signals that are chirping enough, with reconstruction errors that can be reduced by consistently including the corrections we derived. By contrast, low-mass binary black holes, if far away from merger and slowly-chirping, cannot be handled by our perturbative formalism and we develop another approach for these systems. In the case of precessing binaries, we perform a limited exploration of the merger-ringdown range based on a toy model for the precession. We find that, depending on the post-merger frame velocity, deviations from leading order can become qualitatively large to the point of challenging our perturbative formalism, while affecting mainly subdominant mode contributions and giving rise to a limited unfaithfulness in the full waveform. Including higher-order corrections consistently reduces the unfaithfulness, and we further develop an alternative approach to accurately represent post-merger features.

@@

The GW community is making progress assembling higher-fidelity tools for these coming challenges. State-of-the-art IMR templates combine information from the perturbative results of post-Newtonian (PN) theory covering the inspiral (see e.g.~\cite{BlanchetLiving}) and from numerical relativity (NR) simulations covering the end of the inspiral and the merger-ringdown phase (see e.g.~\cite{Pfeiffer12}).  Approaches to template construction include phenomenological templates postulating an analytic ansatz for the Fourier-domain amplitude and phase~\cite{Husa+15, Khan+15,Hannam+13}, and the Effective-One-Body (EOB) approach~\cite{BD99,Taracchini+13, Pan+13, Bohe+16} incorporating PN and NR information. Where needed, Reduced Order Models (ROM), also called surrogate models, have been developed to considerably speed up waveform generation, without losing accuracy~\cite{Field+13, Puerrer14, Blackman+17a}. Put together, these tools provide efficient non-precessing IMR Fourier-domain waveforms, while recent progress has been made for precessing systems as well~\cite{Hannam+13, Chatziioannou+17, Blackman+17b}. Importantly for this work, the resulting waveforms can be represented by an amplitude and phase for each mode, with only a few hundred samples~\cite{Puerrer14}.

%Our challenge will be to accurately complete the representation of instrumental signals including precession effects and the LISA instrumental response and precession effects without sacrificing the computational efficiency enabled by these fast waveform templates. \SM{[rewrite that]}

A complete representation of spin effects in fast IMR templates still remains a frontier of gravitational wave signal modelling. In presence of misaligned spins, the system will endure precession of the orbital plane as it evolves, leading to modulations of the signal as seen by the observer~\cite{Apostolatos+94, Kidder95}. The presence of six degrees of freedom for the spins increases the dimensionality of the problem.

A promising approach to modeling the effect of precession on the emitted waveform, as proposed in~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11}, is to decompose precessing waveforms by performing a time-dependent rotation, following the precession of the orbital plane. The resulting waveform in the rotated frame can then be modelled by a non-precessing waveform, an approximation which is used both in the construction of the inspiral part of precessing EOB waveforms~\cite{Pan+13} and in the construction of precessing phenomenological waveforms~\cite{Hannam+13}. To follow this modelling approach and efficiently create Fourier-domain waveforms, one needs to understand how to translate the time-domain modulations created by the frame rotation into a Fourier-domain transfer function.

Beyond a fast representation of the incident gravitational wave, observational analyses also require transforming the signals through some instrumental response. For short duration mergers such as LIGO  has detected, this can be treated by a simple multiplier and a fixed timeshift between detectors.  For future instruments though, the instrumental response will be more complicated.

@@

Whereas LIGO and Virgo are typically sensitive to chirping binaries for a minute or less, LISA signals may accumulating over months or years. The response of a LISA-type instrument is thus time-dependent~\cite{Cutler97}. The motion and change of orientation of the detector constellation along its orbit lead to significant time variability in the form of a modulation and a varying delay. These effects then convey information about the localization of the gravitational-wave source in the sky. Direct time-domain implementation of the detector response is straightforward~\cite{Vallisneri04, Petiteau+08, CR02, RCP04}, but at a prohibitive computational cost for parameter-estimation analyses. To leverage the performance of state-of-the-art Fourier-domain IMR templates~\cite{BTB16,Khan+15}, we must efficiently process the signals through the time-dependent response of the dectector while staying in the Fourier domain.

@@

In the case of precessing binaries, applying the SPA directly to the modulated signal is also prone to pathologies. In Refs.~\cite{KCY13,KCY14}, a formalism (called shifted uniform asymptotics or SUA) was introduced to go beyond the SPA and compute more accurately the modulation in the Fourier domain; however, this formalism still relies on the SPA for the underlying precessing-frame signal, and is as such limited to inspiraling signals. The simplified treatment of the precession response in the phenomenological waveforms of Ref.~\cite{Hannam+13} takes another approach, treating the precession modulation in the frequency domain by directly associating the Fourier frequency with the post-Newtonain orbital frequency. As we will explain, this corresponds to the zeroth-order approximation of the SUA.

In the case of the LISA response, the SPA provides a natural map from time-domain (for the orbit) to Fourier-domain (for the signal)~\cite{Cutler97}, and was used by many previous studies with inspiral waveforms. Ref.~\cite{Klein+15} included the orbital motion of the detector in the SUA treatment of precessing inspiral signals, within a low-frequency approximation of the constellation response. Consistently extending these previous approaches to the merger-ringdown part of the signals, while including the delays in the full LISA response at all frequencies, is part of the objectives of this paper.

We seek to overcome two limitations in such previous approaches.  The first is that  SPA-based methods are not applicable to IMR waveforms. Second, there is often no clear way to improve the accuracy of these methods beyond the intuitive leading order treatment in order to meet the high-accuracy needs of future detectors. Our approach exploits separation of time-scales approximations, based on a general treatment directly in the Fourier domain of slowly varying delays and amplitude modulations for chirping waveforms.

The plan of the paper is as follows. In Sec.~\ref{sec:motivation}, we provide a general presentation of the problem of Fourier-domain modulation and introduce the relevant timescales for both the response of LISA-type detectors and the modulation of precessing signals. In Sec.~\ref{sec:formalism}, we present our general formalism, give its leading order approximation as well as higher-order corrections, introduce new timescales based on the Fourier-domain signal, refine the previous results for both the quadratic-in-phase corrections and the treatment of the delays, and make the connection to previous approaches of the problem. We then apply our formalism to the response of the LISA detector in Sec.~\ref{sec:LISA}, and to the case of signals from precessing binaries in Sec~\ref{sec:precession}. We discuss and summarize our results in Sec.~\ref{sec:discussion}.

@@

To understand the separation of timescales in the problem, it will be useful to have at hand the leading-order scaling laws for an inspiral (labeled the Newtonian order in the PN language). Although inaccurate for the purpose of waveform modelling, these leading-order estimates will give useful orders of magnitude of the relevant timescales in the inspiral. For a binary with masses $m_{1}, m_{2}$, we define the total mass $M=m_{1}+m_{2}$ and the symmetric mass ratio $\nu = m_{1}m_{2}/M^{2}$. Introducing a time of coalescence $t_{c}$, the relations between the orbital frequency and phase and the time to coalescence $t_{c} - t$ are then given by
\begin{subequations}\label{eq:omegaphiN}
\begin{align}
	\omega(t) &= \left[ \frac{256\nu}{5c^{5}} (GM)^{5/3} (t_{c}-t) \right]^{-3/8} \,, \\
	\varphi(t) &= -\left[ \frac{c^{3}}{5 G M \nu^{3/5}} (t_{c}-t) \right]^{5/8} \,.
\end{align}
\end{subequations}
As for the leading-order time-domain amplitude of the $22$ mode, we have~\cite{BlanchetLiving}
\be\label{eq:a22N}
	a_{22}^{\rm N} (t) = \frac{2 G M \nu v^{2}}{D c^{2}} \sqrt{\frac{16 \pi}{5}} \,,
\ee
where we set $v = (G M\omega/c^{3})^{1/3}$ and where $D$ is the luminosity distance to the observer. For this Newtonian inspiral, applying the SPA gives for the $22$ mode $\tilde{h}(f) = A_{\rm N}(f)e^{-i\Psi_{\rm N}(f)}$ with
\begin{subequations}\label{eq:SPAN}
\begin{align}
	A_{N}(f) &= \frac{G^{2}M^{2} \pi}{Dc^{5}} \sqrt{\frac{2\nu}{3}} v^{-7/2}\,, \label{eq:ASPAN}\\
	\Psi_{\rm SPA}^{\rm N}(f) &= \phi_{0} - 2\pi f t_{0} - \frac{3}{128\nu v^{5}} \,, \label{eq:PsiSPAN} 
\end{align}
\end{subequations}
where $v=(G M \pi f/c^{3})^{1/3}$ according to the SPA correspondence~\eqref{eq:deftfSPA}, and where $t_{0}, \phi_{0}$ are constants. It is also customary to rewrite the above relations in terms of the chirp mass $\Mchirp \equiv M\nu^{3/5}$, which is the only mass combination characterizing the signal at the leading PN order.

@@

@@

An extension is required in specifying a model for the precession post-merger. After the merger occurs, the intuitive interpretation of the precessing frame as roughly following the orbital plane of the binary is lost. However, the prescription of~\cite{OShaughnessy+11} to extract the radiation axis is based entirely on the waveform itself, and can be applied to the ringdown as well. For $\ell=2$, analyzing numerical relativity waveforms, Ref.~\cite{OShaughnessy+12} described a qualitative model where the radiation axis essentially keeps precessing around the final angular momentum $\bm{J}$, but transitions to a faster precession rate, with an angular velocity determined by the quasinormal mode (QNM) frequencies of the remnant black hole as
\be\label{eq:OmegaframeQNM}
	\Omega_{\rm frame} = \omega_{220}^{\rm QNM} - \omega_{210}^{\rm QNM} \,.
\ee
This implies a significant acceleration of the precession post-merger with respect to the inspiral (see Fig.~\ref{fig:precmodel}). For the opening angle of the precessing cone $\beta$, Ref.~\cite{OShaughnessy+12} proposes an exponential decay driven similarly by the difference between damping frequencies. We found by inspection of various precessing waveforms currently available in the SXS catalog (see e.g.~\cite{SXScatalog, Mroue+12, Mroue+13}) that this picture is at least qualitatively correct for $\ell=2$, with $\beta$ appearing to transition to a lower value rather than decaying all the way to zero. By contrast, we found that the current ringdown prescription in the SEOBNRv3 model can cause a qualitatively different behaviour, with oscillations in $\beta$.

@@

Here, we will use directly the wave polarizations $h_{+}$, $h_{\times}$, optimizing over time, phase, and polarization angle. For real or complex signals $a$, $b$, one introduces the usual Hermitian noise-weighted scalar product~\cite{CF94}
\be\label{eq:defoverlap}
	\left( a | b \right) = 2 \int \ud f \frac{\tilde{a}_{1}(f) \tilde{b}_{2}(f)^{*}}{S_{n}(f)} \,,
\ee
where the integral extends over all frequencies (in practice, two intervals $[-f_{\rm max}, -f_{\rm min}]$ and $[f_{\rm min}, f_{\rm max}]$). In the above, $S_{n}(f)$ the noise power spectral density, for which we use the aLIGO ZDHP noise curve~\cite{LIGOProspects13}. We use the notation $h \equiv h_{+} - i h_{\times}$ for the complex strain, and write $h[\delta t, \delta \Phi, \delta \psi]$ to represent a signal where we shifted the time of coalescence, the phase of the line of sight to the observer and the polarization angle. The dependencies in time and polarization are given by~\eqref{eq:shifttime} and $h[\delta \psi] = e^{2 i \delta \psi} h$, while the phase $\delta \Phi$ affects differently each waveform mode in~\eqref{eq:defmodes}. In the norm $(h|h)$, only the dependence on $\delta\Phi$ remains. For two signals $h_{1}$ and $h_{2}$, the mismatch between them is then defined as \SM{[check]}
\be\label{eq:defMM}
	\mathrm{MM} \equiv \mathrm{Min}_{\delta t, \delta \varphi, \delta \psi} \left( 1 - \frac{ \mathrm{Re} \left[ (h_{1}[\delta t, \delta \Phi, \delta \psi] | h_{2}) \right]}{\sqrt{(h_{1}|h_{1})[\delta\Phi]}\sqrt{(h_{2}|h_{2})}} \right)
\ee
where an optimization over a shift in time, phase and polarization is performed. This definition is the same as in~\cite{Blackman+17a} (see App.~D there), except that we added the optimization over the phase. The target unfaithfulness for waveform models depends on the application. For ground-based detectors like advanced LIGO and Virgo, it is often set to 1\% or 3\%, while detectors of the next generations, including LISA, will require a lower unfaithfulness due to their higher signal-to-noise ratios.
